---
title: "STAT 331 Final Project"
subtitle: "Part 1"
author: "Charles Jansen, ..."
format: 
  html:
    code-tools: true
    toc: true
    embed-resources: true
    html-table-processing: none
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

## Part 1: Project Proposal + Data

### Data Description

```{r setup}
# Load necessary libraries here
library(tidyverse)
library(knitr)
```

```{r}
# Load data here
murders <- read_csv("murder_per_100000_people.csv")
unemployment <- read_csv("long_term_unemployment_rate_percent.csv")
```

*Data description goes here...*
** VARIABLES ARE:
** LONG TERM EMPLOYMENT RATE (%)**
** and MURDERS PER 100000 PEOPLE**

### Cleaning Murders Data

First, we'll start with the murders dataset. We can take a look to see if there's anything that clearly needs to change:

```{r}
head(murders)
```
The types of all of the columns seem correct: The country is represented as a character vector, and all of the actual murder rates are numerical. It doesn't seem like we'll need any particular cleaning, so we'll go ahead and pivot our data.

We want to pivot every column that isn't the country into two columns: The year, and the murders per 100,000 in that country and that year.

Additionally, we can convert the year to a number, and drop any rows that do not have an observation for the murders per 100,000.

```{r}
murders_clean <- murders |> 
  pivot_longer(cols = !country, names_to = "year", values_to = "murders_per_100000") |> 
  mutate(year = as.numeric(year)) |> 
  filter(!is.na(murders_per_100000))

head(murders_clean)
```

### Cleaning Unemployment Data

Now that the murders data is clean and pivoted, what about the unemployment data?

First, we should take a look at the data:

```{r}
head(unemployment)
```

Similarly to the murders data, looks good from the start! We'll do the exact same thing to this data as we did to the murders data: Pivot all of the observation columns into one year column and one unemployment rate column, then convert year to a number and drop all missing observations.

```{r}
unemployment_clean <- unemployment |> 
  pivot_longer(cols = !country, names_to = "year", values_to = "long_term_unemployment_rate") |> 
  mutate(year = as.numeric(year)) |> 
  filter(!is.na(long_term_unemployment_rate))

head(unemployment_clean)
```

### Joining Datasets

Before we join the